---
title: "CRJ 605 (Statistical Analysis of Networks) Class Survey Data"
date: "Network Description"
output: 
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_float: yes
    code_fold: show
---

```{r, echo=FALSE, eval=TRUE, message=FALSE}
# clear off the workspace.
rm(list = ls())

# Load the data from local.
setwd("/Users/jyoung20/Dropbox (ASU)/TEACHING_materials/ASU_courses/CRJ_605_Networks_Young/CRJ_605_Networks_SPRING_2019/")
load("CRJ_605_Networks_SPRING_2019_Class_Survey_Data_W1.RData")
# Or from my website.
load(url("https://www.jacobtnyoung.com/uploads/2/3/4/5/23459640/crj_605_networks_spring_2019_class_survey_data_w1.rdata"))

# Load the libraries you will need.
library(network)
library(sna)
library(xtable) # this package is for building tables in RMarkdown

```

This file provides some description of the course survey data that was collected the first day of class.

#Description
There are three networks: 

* `spend.time.net` the Spent Time With network
* `talk.course.net` the Talked about the Course network
* `trust.net` the Trust with Damaging information network  

For each network, the survey data have been attached to the objects as vertex attributes. Also, each network object the two other networks have been attached as edge attributes.


Table ` below gives several descriptives for the each of the networks:  

```{r,echo=FALSE,eval=TRUE, message=FALSE, warning=FALSE}
# Descriptives table for the networks.

edges.ST    <- sum(as.matrix(spend.time.net))
edges.TC    <- sum(as.matrix(talk.course.net))
edges.TR    <- sum(as.matrix(trust.net))
recip.ST    <- grecip(spend.time.net,  measure="edgewise")
recip.TC    <- grecip(talk.course.net,  measure="edgewise")
recip.TR    <- grecip(trust.net,  measure="edgewise")
trans.ST    <- gtrans(spend.time.net)
trans.TC    <- gtrans(talk.course.net)
trans.TR    <- gtrans(trust.net)
mean.in.ST  <- mean(degree(spend.time.net, cmode="indegree"))
mean.in.TC  <- mean(degree(talk.course.net, cmode="indegree"))
mean.in.TR  <- mean(degree(trust.net, cmode="indegree"))
sd.in.ST    <- sd(degree(spend.time.net, cmode="indegree"))
sd.in.TC    <- sd(degree(talk.course.net, cmode="indegree"))
sd.in.TR    <- sd(degree(trust.net, cmode="indegree"))
mean.out.ST <- mean(degree(spend.time.net, cmode="outdegree"))
mean.out.TC <- mean(degree(talk.course.net, cmode="outdegree"))
mean.out.TR <- mean(degree(trust.net, cmode="outdegree"))
sd.out.ST   <- sd(degree(spend.time.net, cmode="outdegree"))
sd.out.TC   <- sd(degree(talk.course.net, cmode="outdegree"))
sd.out.TR   <- sd(degree(trust.net, cmode="outdegree"))

ties          <- rbind(edges.ST,edges.TC,edges.TR)
recip         <- rbind(recip.ST,recip.TC,recip.TR)
trans         <- rbind(trans.ST,trans.TC,trans.TR)
mean.in       <- rbind(mean.in.ST,mean.in.TC,mean.in.TR)
sd.in         <- rbind(sd.in.ST,sd.in.TC,sd.in.TR)
mean.out      <- rbind(mean.out.ST,mean.out.TC,mean.out.TR)
sd.out        <- rbind(sd.out.ST,sd.out.TC,sd.out.TR)
mat           <- cbind(ties,recip,trans,mean.in,sd.in,mean.out,sd.out)
colnames(mat) <- c("Edges","Reciprocity","Transitivity","Mean Indgree", "SD Indegree", "Mean Outdegree", "SD Outdegree")
rownames(mat) <- c("Spend Time With (ST)", "Talked about Course (TC)", "Trust (TR)")
tab           <- xtable(mat, caption = "Table 1: Descriptive Statistics for CRJ 605 (Spring 2019) Course Survey Data Networks", digits=c(0,0,2,2,2,2,2,2))
```    


```{r xtable1, results="asis", echo=FALSE}
#Print the table.
print.xtable(tab, type="html", caption.placement = "top")
```  

***


