---
title: "Lab 04-Degree Centrality and Centralization"
date: "CRJ 605 Statistical Analysis of Networks"
output: 
  html_document:
    df_print: paged
    theme: lumen
    highlight: haddock
    toc: yes
    toc_float: yes
    code_fold: show
    self_contained: true
---

```{r,echo=FALSE,eval=TRUE,message=FALSE}

library( sna )
library( devtools )
library( UserNetR )
library( network )

```

<br>

*This lab examines degree centrality and centralization using the `degree()` and `centralization()` functions in the `sna` package.*

<br>

***

### **Degree Centrality (Undirected Binary Graphs)**  

In an undirected binary graph, *actor degree centrality* measures the extent to which a node connects to all other nodes in a network. In other words, the number of edges incident with a node. This is symbolized as: $d(n_i)$. For an undirected binary graph, the degree $d(n_i)$ is the row or column sum. If we have an object of `class(matrix)` in the workspace, we can use the `colSums()` and/or `rowSums()` functions to return this information.  

First, let's set up our graph from the degree centrality lecture:  

```{r,echo=TRUE,eval=TRUE,message=FALSE}

# First, clear the workspace
rm( list = ls() )

# Then, build an object
u.mat <- rbind(
  c( 0,1,0,0,0 ),
  c( 1,0,1,0,0 ),
  c( 0,1,0,1,1 ),
  c( 0,0,1,0,1 ),
  c( 0,0,1,1,0 ) )

# Assign the names to the object
rownames( u.mat ) <- c( "Jen","Tom","Bob","Leaf","Jim" )
colnames( u.mat ) <- c( "Jen","Tom","Bob","Leaf","Jim" )

# Now, plot the graph (remember to load the sna package)
# The quitely= argument tells R not to print out the info on the package
library( sna, quietly=TRUE ) 

# Let's set up the coordinates to force the nodes
# to be in the same position throughout the lab
set.seed( 605 )
coords <- gplot( u.mat )

# Plot the network
gplot( 
  u.mat, 
  gmode="graph", 
  arrowhead.cex=0.5, 
  edge.col="grey40", 
  label=rownames( u.mat ),
  label.col="blue",
  label.cex=1.2,
  coord = coords
  )

```  

Since the graph is undirected, we can print the degree centrality for each node as a vector using the `colSums()` or `rowSums()` functions:  

```{r,echo=TRUE,eval=TRUE,message=FALSE}

colSums( u.mat )
rowSums( u.mat )

# We could also assign these to an object
deg.u.mat <- colSums( u.mat )

# Then, use that information in the plot
# Use the vertex.cex= argument to pass the degree
# This will make nodes with higher degree larger
gplot(
  u.mat,
  gmode="graph", 
  arrowhead.cex=0.5, 
  edge.col="grey40", 
  label=rownames( u.mat ),
  label.col="blue",
  label.cex=1.2,
  vertex.cex = deg.u.mat,
  coord = coords
  )

# Or, we could use the RColorBrewer package to shade the nodes

# install.packages( "RColorBrewer" )
library( RColorBrewer, quietly=TRUE )

# use display.brewer.all() to see the pallettes.

# Let's use the Blues pallette.
col.deg  <- brewer.pal( length( unique( deg.u.mat ) ), "Blues")[deg.u.mat]

# In this plot, what do darker shades mean?
gplot(
  u.mat, 
  gmode="graph", 
  arrowhead.cex=0.5, 
  edge.col="grey40", 
  label=rownames( u.mat ),
  label.col="blue",
  label.cex=1.2,
  vertex.cex = deg.u.mat,
  vertex.col = col.deg,
  coord = coords
  )

```  

#### **Standardized degree centrality, mean degree, and centralization**  
Actor degree centrality not only reflects each node’s connectivity to other nodes but also depends on the size of the network, *g*. As a result, larger networks will have a higher maximum possible degree centrality values. This makes comparison across networks problematic. The solution is to take into account the number of nodes and the maximum possible nodes to which *i* could be connected, *g-1*.  

Let's calculate the standardized centrality scores for our undirected graph:  

```{r,echo=TRUE,eval=TRUE,message=FALSE}

# unstandardized or raw centrality
deg.u.mat <- colSums( u.mat )

# to calculate g-1, we need to know the number of nodes in the graph 
# this is the first dimension of the matrix
g <- dim( u.mat )[1]

# now, divide by g-1
s.deg.u.mat <- deg.u.mat / ( g-1 )

deg.u.mat
s.deg.u.mat

```

We can also examine the *average* degree of the graph using 

$$\frac{\sum_{i=1}^g d(n_i)}{g}$$
or 

$$\frac{2L}{g}$$, 

where *L* is the number of edges in the graph:

```{r,echo=TRUE,eval=TRUE,message=FALSE}

mean.deg <- sum( deg.u.mat ) / dim( u.mat )[1] 

mean.deg

# Note that we can also use the mean() function to return this information:
mean( deg.u.mat )

```  

<br>

We can also calculate how centralized the graph itself is. *Group degree centralization* measures the extent to which the actors in a social network differ from one another in their individual degree centralities. Following [Wasserman & Faust (1994)](https://www.cambridge.org/core/books/social-network-analysis/90030086891EB3491D096034684EFFB8), an index of *group degree centralization* can be calculated as: 

$$C_D = \frac{\sum\limits_{i=1}^g [C_D(n^*) - C_D(n_i)]}{[(g-1)(g-2)]}$$

for undirected graphs where $C_D(n^*)$ is the maximum degree in the graph. We can write out the components of the equation using the `max()` function:

```{r,echo=TRUE,eval=TRUE,message=FALSE}

# In separate pieces
deviations <- max( deg.u.mat ) - deg.u.mat
sum.deviations <- sum( deviations )
numerator <- sum.deviations
denominator <- ( g-1 )*( g-2 )
group.deg.cent <- numerator/denominator

group.deg.cent


# Or, as a single equation.
group.deg.cent <-( sum( ( ( max( deg.u.mat ) - deg.u.mat ) ) ) ) / ( ( g -1 )*( g - 2 ) )

group.deg.cent
 
```  

***

### **Degree Centrality (Directed Binary Graphs)**  

In a directed binary graph, actor degree centrality can be broken down into *indegree* and *outdegree* centrality. Indegree, $C_I(n_i)$, measures the number of ties that *i* receives. For the sociomatrix $Xij$, the indegree for *i* is the **column** sum. Outdegree, $C_O(n_i)$, measures the number of ties that *i* sends. For the sociomatrix $Xij$, the outdegree for *i* is the **row** sum.   

As before, if we have an object of `class(matrix)` in the workspace, we can use the `rowSums()` and `colSums()` functions. However, the `colSums()` function will return the *indegree* centrality for *i* and the `rowSums()` function will return the *outdegree* centrality for *i*.    

First, let's set up our directed graph from the degree centrality lecture:  

```{r,echo=TRUE,eval=TRUE,message=FALSE,warning=FALSE}

# First, clear the workspace
rm( list = ls() )

# Then, build the object
d.mat <- rbind(
  c( 0,1,0,0,0 ),
  c( 0,0,1,0,0 ),
  c( 0,0,0,1,1 ),
  c( 0,0,1,0,1 ),
  c( 0,0,1,1,0 ) )
rownames( d.mat ) <- c( "Jen","Tom","Bob","Leaf","Jim" )
colnames( d.mat ) <- c( "Jen","Tom","Bob","Leaf","Jim" )

# Let's set up the coordinates to force the nodes
# to be in the same position throughout the lab
set.seed( 605 )

# remove the old object named coords
rm( coords )

# set the new coordinates
coords <- gplot( d.mat )

# Now, plot the graph (remember to load the sna package)
gplot(
  d.mat, 
  gmode="digraph",
  arrowhead.cex=0.5, 
  edge.col="grey40", 
  label=rownames( d.mat ),
  label.col="red",
  label.cex=1.2,
  coord = coords
  )

```

```{r,echo=TRUE,eval=TRUE,message=FALSE,warning=FALSE, fig.align="center"}

# Let's look at the different centrality scores 
# by assigning them to different objects
ideg.d.mat <- colSums( d.mat )
odeg.d.mat <- rowSums( d.mat )

# Then, use that information in the plot
# partition the plotting window using 
# the par() function to show two plots
# change the margins using the mar= argument
par( 
  mfrow=c( 1, 2 ), 
  mar=c( 0.1, 0.5, 1, 0.1) 
  )

gplot(
  d.mat, 
  gmode="digraph", 
  arrowhead.cex=0.5, 
  edge.col="grey40", 
  label=rownames( d.mat ),
  label.col="red",
  label.cex=0.8,
  label.pos=1,
  vertex.cex = ideg.d.mat+0.2,
  main="Nodes sized by Indegree",
  coord = coords
  )

gplot(
  d.mat, 
  gmode="digraph", 
  arrowhead.cex=0.5, 
  edge.col="grey40", 
  label=rownames( d.mat ),
  label.col="red",
  label.cex=0.8,
  label.pos=1,
  vertex.cex = odeg.d.mat,
  main="Nodes sized by Outdegree",
  coord = coords
  )

# return plot partition to 1 pane
par( mfrow=c( 1,1 ) )

```

```{r,echo=TRUE,eval=TRUE,message=FALSE,warning=FALSE}

# Or, we could use the RColorBrewer package to shade the nodes
library( RColorBrewer, quietly=TRUE )


# create the objects
col.ideg  <- brewer.pal( length( unique( ideg.d.mat ) ), "Greens")[ideg.d.mat]
col.odeg  <- brewer.pal( length( unique( odeg.d.mat ) ), "Oranges")[odeg.d.mat]


par( 
  mfrow=c( 1, 2 ), 
  mar=c( 0.1, 0.5, 1, 0.1) 
  )

gplot(
  d.mat, 
  gmode = "digraph", 
  arrowhead.cex = 0.5, 
  edge.col = "grey40", 
  label = rownames( d.mat ),
  label.col = "red",
  label.cex = 0.8,
  label.pos = 1,
  vertex.cex = ideg.d.mat+0.2,
  vertex.col = col.ideg,
  main = "Nodes sized & shaded by Indegree",
  coord = coords
  )

gplot(
  d.mat, 
  gmode = "digraph", 
  arrowhead.cex = 0.5, 
  edge.col = "grey40", 
  label = rownames( d.mat ),
  label.col = "red",
  label.cex = 0.8,
  label.pos = 1,
  vertex.cex = odeg.d.mat,
  vertex.col = col.odeg,
  main = "Nodes sized & shaded by Outdegree",
  coord = coords
  )

# return plot partition to 1 pane
par( mfrow=c( 1,1 ) )

```

<br>

#### **Standardized degree centrality, mean degree, and centralization**  

Let's calculate the standardized centrality scores for our directed graph:  

```{r,echo=TRUE,eval=TRUE,message=FALSE}

# unstandardized or raw centrality
ideg.d.mat <- colSums( d.mat )
odeg.d.mat <- rowSums( d.mat )

# to calculate g-1, we need to know the number of nodes in the graph
# this is the first dimension of the matrix
g <- dim( d.mat )[1]

# now, divide by g-1
s.i.deg.u.mat <- ideg.d.mat / ( g-1 )
s.o.deg.u.mat <- odeg.d.mat / ( g-1 )
```

We can also examine the *average* degree of the graph using $\frac{\sum_{i=1}^g C_I(n_i)}{g} = \frac{\sum_{i=1}^g C_O(n_i)}{g}$ or $\frac{L}{g}$, where *L* is the number of edges in the graph:

```{r,echo=TRUE,eval=TRUE,message=FALSE}

mean.i.deg <- sum( ideg.d.mat ) / dim( d.mat )[1] 
mean.o.deg <- sum( odeg.d.mat ) / dim( d.mat )[1] 
mean.i.deg
mean.o.deg

```  

Again, following Wasserman & Faust (1994), an index of *group indegree/outdegree centralization* can be calculated as:  

  $$ C_D = \frac{\sum\limits_{i=1}^g [C_D(n^*) - C_D(n_i)]}{[(g-1)^2]} $$  

for undirected graphs where $C_D(n^*)$ is the maximum indegree/outdegree in the graph. We can write out the components of the equation using the `max()` function:

```{r,echo=TRUE,eval=TRUE,message=FALSE}

# In separate pieces
deviations <- max( ideg.d.mat ) - ideg.d.mat
sum.deviations <- sum( deviations )
numerator <- sum.deviations
denominator <- ( g-1 )*( g-1 )
group.i.deg.cent <- numerator/denominator
group.i.deg.cent

deviations <- max( odeg.d.mat ) - odeg.d.mat
sum.deviations <- sum( deviations )
numerator <- sum.deviations
denominator <- ( g-1 )*( g-1 )
group.o.deg.cent <- numerator/denominator
group.o.deg.cent


# Or, as a single equation
group.ideg.cent <-( sum( ( ( max( ideg.d.mat ) - ideg.d.mat ) ) ) ) / ( ( g -1 )*( g - 1) )
group.odeg.cent <-( sum( ( ( max( odeg.d.mat ) - odeg.d.mat ) ) ) ) / ( ( g -1 )*( g - 1 ) )
group.ideg.cent
group.odeg.cent
```


***  

### **Degree Centrality using the `sna` package**  
Did that feel tedious? If no, go back and do it again :)  

As you probably have guessed, there *are* functions in the `sna` package that calcuates degree centrality and graph centralization! These are the `degree()` and `centralization()` functions, respectively. Let's take a look at how these work.

```{r,echo=TRUE,eval=TRUE,message=FALSE}
 
# load the library
library( sna )

# Build the objects to work with
rm( list = ls() )

u.mat <- rbind( 
  c( 0,1,0,0,0 ),
  c( 1,0,1,0,0 ), 
  c( 0,1,0,1,1 ), 
  c( 0,0,1,0,1 ), 
  c( 0,0,1,1,0 )
  )

rownames( u.mat ) <- c( "Jen","Tom","Bob","Leaf","Jim" )

colnames( u.mat ) <- c( "Jen","Tom","Bob","Leaf","Jim" )

d.mat <- rbind(
  c( 0,1,0,0,0 ),
  c( 0,0,1,0,0 ), 
  c( 0,0,0,1,1 ), 
  c( 0,0,1,0,1 ), 
  c( 0,0,1,1,0 )
  )

rownames( d.mat ) <- c( "Jen","Tom","Bob","Leaf","Jim" )

colnames( d.mat ) <- c( "Jen","Tom","Bob","Leaf","Jim" )


# First, let's look at degree
?degree

# degree for undirected graph
deg <- degree( u.mat, gmode="graph" )

# indegree for directed graph
ideg <- degree( d.mat, gmode="digraph", cmode="indegree" )

# outdegree for directed graph
odeg <- degree( d.mat, gmode="digraph", cmode="outdegree" )

# returns the combined centrality for each node
deg.d <- degree( d.mat, gmode="digraph" )


# Now, let's look at centralization
?centralization

# degree centralization for undirected graph
cent.u <- centralization( u.mat, degree, mode="graph" )

# indegree centralization for directed graph.
i.cent.d <- centralization( d.mat, degree, mode="digraph", cmode="indegree" ) 

# outdegree centralization for directed graph.
o.cent.d <- centralization( d.mat, degree, mode="digraph", cmode="outdegree" )

```  

<br>

*Now, wasn't that easier?*

<br>

***  

#### ***Questions?***


<br>

##### ***Please*** report any needed corrections to the [Issues](https://github.com/jacobtnyoung/SAND/issues/new) page. Thanks!

<br>
<br>


###### ***Last updated `r format(Sys.time(), '%d %B, %Y')`***