---
title: "Degree Centrality and Centralization in R Lab"
date: "CRJ 605 Statistical Analysis of Networks"
output: 
  html_document:
    df_print: paged
    theme: cosmo
    highlight: haddock
    toc: yes
    toc_float: yes
    code_fold: show
---

```{r,echo=FALSE,eval=TRUE,message=FALSE}
library(sna)
library(devtools)
#library(UserNetR)
library(network)
```

###
*This lab examines degree centrality and centralization using the `degree()` and `centralization()` functions in the `sna` package.*

***

###**Degree Centrality (Undirected Binary Graphs)**  

In an undirected binary graph, *actor degree centrality* measures the extent to which a node connects to all other nodes in a network. In other words, the number of edges incident with a node. This is symbolized as: $d(n_i)$. For an undirected binary graph, the degree $d(n_i)$ is the row or column sum. If we have an object of `class(matrix)` in the workspace, we can use the `colSums()` and/or `rowSums()` functions to return this information.  

First, let's set up our graph from the degree centrality lecture:  

```{r,echo=TRUE,eval=TRUE,message=FALSE}
# First, clear the workspace.
rm(list = ls())

# Then, build the object.
u.mat <- rbind(
  c(0,1,0,0,0),
  c(1,0,1,0,0),
  c(0,1,0,1,1),
  c(0,0,1,0,1),
  c(0,0,1,1,0))
rownames(u.mat) <- c("Jen","Tom","Bob","Leaf","Jim")
colnames(u.mat) <- c("Jen","Tom","Bob","Leaf","Jim")

# Now, plot the graph (remember to load the sna package).
library(sna, quietly=TRUE) # the quitely argument tells R not to print out the info on the package.
gplot(u.mat, gmode="graph", arrowhead.cex=0.5, edge.col="grey40", label=rownames(u.mat),label.col="Blue",label.cex=1.2)

```  

Since the graph is undirected, we can print the degree centrality for each node as a vector using the `colSums()` or `rowSums()` functions:  

```{r,echo=TRUE,eval=TRUE,message=FALSE}
colSums(u.mat)
rowSums(u.mat)

# We could also assign these to an object.
deg.u.mat <- colSums(u.mat)

# Then, use that information in the plot.
gplot(u.mat, gmode="graph", arrowhead.cex=0.5, edge.col="grey40", 
      label=rownames(u.mat),label.col="Blue",label.cex=1.2,
      vertex.cex = deg.u.mat #here, larger nodes have higher degree. 
      )

# Or, we could use the RColorBrewer package to shade the nodes.
#install.packages("RColorBrewer")
library(RColorBrewer, quietly=TRUE)
  # use display.brewer.all() to see the pallettes.

# Let's use the Blues pallette.
col.deg  <- brewer.pal(length(unique(deg.u.mat)), "Blues")[deg.u.mat]
gplot(u.mat, gmode="graph", arrowhead.cex=0.5, edge.col="grey40", 
      label=rownames(u.mat),label.col="Blue",label.cex=1.2,
      vertex.cex = deg.u.mat, #here, larger nodes have higher degree.
      vertex.col = col.deg #here, bluer (?) nodes have higher degree. 
      )
```  

####**Standardized degree centrality, mean degree, and centralization**  
Actor degree centrality not only reflects each node’s connectivity to other nodes but also depends on the size of the network, *g*. As a result, larger networks will have a higher maximum possible degree centrality values. This makes comparison across networks problematic. The solution is to take into account the number of nodes and the maximum possible nodes to which *i* could be connected, *g-1*.  

Let's calculate the standardized centrality scores for our undirected graph:  

```{r,echo=TRUE,eval=TRUE,message=FALSE}
# unstandardized or raw centrality.
deg.u.mat <- colSums(u.mat)

# to calculate g-1, we need to know the number of nodes in the graph. 
# this is the first dimension of the matrix.
g <- dim(u.mat)[1]

# now, divide by g-1.
s.deg.u.mat <- deg.u.mat / (g-1)
```

We can also examine the *average* degree of the graph using $\frac{\sum_{i=1}^g d(n_i)}{g}$ or $\frac{2L}{g}$, where *L* is the number of edges in the graph:
```{r,echo=TRUE,eval=TRUE,message=FALSE}
mean.deg <- sum(deg.u.mat) / dim(u.mat)[1] 
mean.deg

#Note that we can also use the mean() function to return this information:
mean(deg.u.mat)

```  

We can also calculate how centralized the graph itself is. *Group degree centralization* measures the extent to which the actors in a social network differ from one another in their individual degree centralities. Following Wasserman & Faust (1994), an index of *group degree centralization* can be calculated as: $C_D = \frac{\sum\limits_{i=1}^g [C_D(n^*) - C_D(n_i)]}{[(g-1)(g-2)]}$ for undirected graphs where $C_D(n^*)$ is the maximum degree in the graph. We can write out the components of the equation using the `max()` function:

```{r,echo=TRUE,eval=TRUE,message=FALSE}

# In separate pieces.
deviations <- max(deg.u.mat) - deg.u.mat
sum.deviations <- sum(deviations)
numerator <- sum.deviations
denominator <- (g-1)*(g-2)
group.deg.cent <- numerator/denominator
group.deg.cent

# Or, as a single equation.
group.deg.cent <-(sum(((max(deg.u.mat) - deg.u.mat)))) / ((g -1)*(g - 2))
group.deg.cent
 
```  

***

###**Degree Centrality (Directed Binary Graphs)**  

In a directed binary graph, actor degree centrality can be broken down into *indegree* and *outdegree* centrality. Indegree, $C_I(n_i)$, measures the number of ties that *i* receives. For the sociomatrix $Xij$, the indegree for *i* is the **column** sum. Outdegree, $C_O(n_i)$, measures the number of ties that *i* sends. For the sociomatrix $Xij$, the outdegree for *i* is the **row** sum.   

As before, if we have an object of `class(matrix)` in the workspace, we can use the `rowSums()` and `colSums()` functions. However, the `colSums()` function will return the *indegree* centrality for *i* and the `rowSums()` function will return the *outdegree* centrality for *i*.    

First, let's set up our directed graph from the degree centrality lecture:  

```{r,echo=TRUE,eval=TRUE,message=FALSE,warning=FALSE}
# First, clear the workspace.
rm(list = ls())

# Then, build the object.
d.mat <- rbind(
  c(0,1,0,0,0),
  c(0,0,1,0,0),
  c(0,0,0,1,1),
  c(0,0,1,0,1),
  c(0,0,1,1,0))
rownames(d.mat) <- c("Jen","Tom","Bob","Leaf","Jim")
colnames(d.mat) <- c("Jen","Tom","Bob","Leaf","Jim")

# Now, plot the graph (remember to load the sna package).
gplot(d.mat, gmode="digraph", #note that we have toggled the gmode= argument to digraph, which is the input for a directed graph.
      arrowhead.cex=0.5, edge.col="grey40", label=rownames(d.mat),label.col="Red",label.cex=1.2
      )

# Let's look at the different centrality scores by assigning them to different objects.
ideg.d.mat <- colSums(d.mat)
odeg.d.mat <- rowSums(d.mat)


# Then, use that information in the plot.
par(mfrow=c(2,1)) # partition the plotting window to show two plots.

gplot(d.mat, gmode="digraph", arrowhead.cex=0.5, edge.col="grey40", 
      label=rownames(d.mat),label.col="Red",label.cex=0.8,
      vertex.cex = ideg.d.mat+0.2, #here, larger nodes have higher indegree. 
      sub="Nodes sized by Indegree"
      )
gplot(d.mat, gmode="digraph", arrowhead.cex=0.5, edge.col="grey40", 
      label=rownames(d.mat),label.col="Red",label.cex=0.8,
      vertex.cex = odeg.d.mat, #here, larger nodes have higher indegree. 
      sub="Nodes sized by Outdegree"
      )
par(mfrow=c(1,1)) # return plot partition to 1 pane.

# Or, we could use the RColorBrewer package to shade the nodes.
#install.packages("RColorBrewer")
library(RColorBrewer, quietly=TRUE)
  # use display.brewer.all() to see the pallettes.
col.ideg  <- brewer.pal(length(unique(ideg.d.mat)), "Greens")[ideg.d.mat]
col.odeg  <- brewer.pal(length(unique(odeg.d.mat)), "Oranges")[odeg.d.mat]

par(mfrow=c(2,1)) # partition the plotting window to show two plots.
gplot(d.mat, gmode="digraph", arrowhead.cex=0.5, edge.col="grey40", 
      label=rownames(d.mat),label.col="Green",label.cex=0.8,
      vertex.cex = ideg.d.mat+0.2, #here, larger nodes have higher indegree.
      vertex.col = col.ideg, #here, greener nodes have higher indegree. 
      sub="Nodes sized & shadded by Indegree"
      )
gplot(d.mat, gmode="digraph", arrowhead.cex=0.5, edge.col="grey40", 
      label=rownames(d.mat),label.col="Orange",label.cex=0.8,
      vertex.cex = odeg.d.mat+0.2, #here, larger nodes have higher outdegree.
      vertex.col = col.odeg, #here, greener nodes have higher outdegree. 
      sub="Nodes sized & shadded by Outdegree"
      )
par(mfrow=c(1,1)) # return plot partition to 1 pane.
```  

####**Standardized degree centrality, mean degree, and centralization**  
Let's calculate the standardized centrality scores for our directed graph:  

```{r,echo=TRUE,eval=TRUE,message=FALSE}
# unstandardized or raw centrality.
ideg.d.mat <- colSums(d.mat)
odeg.d.mat <- rowSums(d.mat)

# to calculate g-1, we need to know the number of nodes in the graph. 
# this is the first dimension of the matrix.
g <- dim(d.mat)[1]

# now, divide by g-1.
s.i.deg.u.mat <- ideg.d.mat / (g-1)
s.o.deg.u.mat <- odeg.d.mat / (g-1)
```

We can also examine the *average* degree of the graph using $\frac{\sum_{i=1}^g C_I(n_i)}{g} = \frac{\sum_{i=1}^g C_O(n_i)}{g}$ or $\frac{L}{g}$, where *L* is the number of edges in the graph:
```{r,echo=TRUE,eval=TRUE,message=FALSE}
mean.i.deg <- sum(ideg.d.mat) / dim(d.mat)[1] 
mean.o.deg <- sum(odeg.d.mat) / dim(d.mat)[1] 
mean.i.deg
mean.o.deg
```  

Again, following Wasserman & Faust (1994), an index of *group indegree/outdegree centralization* can be calculated as: $C_D = \frac{\sum\limits_{i=1}^g [C_D(n^*) - C_D(n_i)]}{[(g-1)^2]}$ for undirected graphs where $C_D(n^*)$ is the maximum indegree/outdegree in the graph. We can write out the components of the equation using the `max()` function:

```{r,echo=TRUE,eval=TRUE,message=FALSE}

# In separate pieces.
deviations <- max(ideg.d.mat) - ideg.d.mat
sum.deviations <- sum(deviations)
numerator <- sum.deviations
denominator <- (g-1)*(g-1)
group.i.deg.cent <- numerator/denominator
group.i.deg.cent

deviations <- max(odeg.d.mat) - odeg.d.mat
sum.deviations <- sum(deviations)
numerator <- sum.deviations
denominator <- (g-1)*(g-1)
group.o.deg.cent <- numerator/denominator
group.o.deg.cent


# Or, as a single equation.
group.ideg.cent <-(sum(((max(ideg.d.mat) - ideg.d.mat)))) / ((g -1)*(g - 1))
group.odeg.cent <-(sum(((max(odeg.d.mat) - odeg.d.mat)))) / ((g -1)*(g - 1))
group.ideg.cent
group.odeg.cent
```


***  

###**Degree Centrality using the `sna` package**  
Did that feel tedious? If no, go back and do it again :)  

As you probably have guessed, there *are* functions in the `sna` package that calcuates degree centrality and graph centralization! These are the `degree()` and `centralization()` functions, respectively. Let's take a look at how these work.

```{r,echo=TRUE,eval=TRUE,message=FALSE}
library(sna) #load the library.

# Build the objects to work with.
rm(list = ls())
u.mat <- rbind(c(0,1,0,0,0),c(1,0,1,0,0), c(0,1,0,1,1), c(0,0,1,0,1), c(0,0,1,1,0))
rownames(u.mat) <- c("Jen","Tom","Bob","Leaf","Jim")
colnames(u.mat) <- c("Jen","Tom","Bob","Leaf","Jim")
d.mat <- rbind(c(0,1,0,0,0),c(0,0,1,0,0), c(0,0,0,1,1), c(0,0,1,0,1), c(0,0,1,1,0))
rownames(d.mat) <- c("Jen","Tom","Bob","Leaf","Jim")
colnames(d.mat) <- c("Jen","Tom","Bob","Leaf","Jim")

# First, let's look at degree.
?degree

deg <- degree(u.mat,gmode="graph") #degree for undirected graph.

ideg <- degree(d.mat,gmode="digraph",cmode="indegree") # indegree for directed graph.
odeg <- degree(d.mat,gmode="digraph",cmode="outdegree") # outdegree for directed graph.

deg.d <- degree(d.mat,gmode="digraph") #returns the combined centrality for each node.


# Now, let's look at centralization.
?centralization

cent.u <- centralization(u.mat, degree, mode="graph") # degree centralization for undirected graph.

i.cent.d <- centralization(d.mat, degree, mode="digraph",cmode="indegree") # indegree centralization for directed graph.
o.cent.d <- centralization(d.mat, degree, mode="digraph",cmode="outdegree") # outdegree centralization for directed graph.

```  
*Now, wasn't that easier?*


***  

####***Questions?***####

```{r,echo=FALSE,eval=FALSE}
#END OF LAB
```