---
title: "Homework 2 - Network Data Structures & Network Visualization"
author: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    df_print: paged
    theme: journal
    highlight: haddock
    toc: yes
    toc_float: yes
    code_fold: show
    self_contained: true
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 10, 
                      fig.height = 6)

```

<br>

```{r load_packages}
# make sure these are installed on your machine before knitting
# install.packages( "network" )
# install.packages( "sna" )
#
# Note that you cannot knit with an install.packages() function
# called in your RMD file. You only have to install packages
# once, then they are loaded using the library() function.

library( network )
library( sna )

```

<br>

As indicated in the instructions, for *each* network you created, do the following:  

1. Create an adjacency matrix of the network and save it as a .csv file.
2. Import the network into *R* using the `read.csv()` function or create the network using the `matrix()` function.
3. Create an object of class `network`. 
4. Plot the network using the `gplot()` function.    
5. Modify the plot using at least three arguments in the `gplot()` function.  

As a refresher, review the example below.

<br>

## Example: 

If you are importing your network, use the `read.csv()` function:

```{r import-example, echo = TRUE, eval = TRUE}

# define the path
url <- "https://github.com/jacobtnyoung/sna-textbook/raw/main/data/data-undirected-example.csv"

# import the undirected matrix
mat.u <- read.csv( 
  url,
  as.is = TRUE,
  header = TRUE,
  row.names = 1 
  )

```

If you are creating your network, use the `matrix()` function:

<br>

```{r create-example, echo = TRUE, eval = TRUE}

# created a directed matrix
mat.d <- matrix(
  c(0,1,0,0,0,1,0,1,0,0,0,1,0,1,1,0,0,1,0,1,0,0,1,1,0),
  nrow=5,
  byrow=TRUE
  )

# attach row names
rownames( mat.d ) <- c( "Jen","Tom","Bob","Leaf","Jim" )

# attach column names
colnames( mat.d ) <- c( "Jen","Tom","Bob","Leaf","Jim" )

# print out the object
mat.d

```

<br>

Finally, let's plot the networks using the `gplot()` function.

```{r create-undirected-plot, echo = TRUE, eval = TRUE}

# produce the plot
gplot(
  mat.u,                          # this is the network to plot
  gmode = "graph",                # let it know it is undirected    
  vertex.col = "lightblue",       # color the nodes
  vertex.cex = 1.2,               # size of the nodes (bigger number is bigger nodes)
  edge.col = "grey80",            # color the ties
  edge.lwd = 1.2,                 # the width of the ties (bigger is wider)
  main = "Plot of undirected net" # add a title
)

```

<br>

```{r create-directed-plot, echo = TRUE, eval = TRUE}

# produce the plot
gplot(
  mat.d,                          # this is the network to plot
  vertex.col = "firebrick",       # color the nodes
  vertex.cex = 1.2,               # size of the nodes (bigger number is bigger nodes)
  edge.col = "grey80",            # color the ties
  edge.lwd = 1.2,                 # the width of the ties (bigger is wider)
  main = "Plot of a directed net" # add a title
)

```

<br>

These were some examples to help you get going. Feel free to use them to aid in working through the homework.

----

# Part I

<br>

## First Network

<br>

### Step 2

Import the network into *R* using the `read.csv()` function or create the network using the `matrix()` function.

```{r}

# Your code here

```

<br>

### Step 3

Create an object of class `network`. 

```{r}

# Your code here

```

<br>

### Steps 4 & 5

Create an object of class `network`. Plot the network using the `gplot()` function. Modify the plot using at least three arguments in the `gplot()` function.    

<br>

----

# Part II

<br>

## Second Network

<br>

### Step 2

Import the network into *R* using the `read.csv()` function or create the network using the `matrix()` function.

```{r}

# Your code here

```

<br>

### Step 3

Create an object of class `network`. 

```{r}

# Your code here

```

<br>

### Steps 4 & 5

Create an object of class `network`. Plot the network using the `gplot()` function. Modify the plot using at least three arguments in the `gplot()` function.    

<br>

```{r}

# Your code here

```

<br>
<br>

<center>------------<h4> **DELETE THIS LINE & ALL LINES BELOW BEFORE SUBMITTING** </h4>------------</center>

<br>
<br>

## Knitting and Submitting

For all the homework assignments, be sure to show both the solution for your code and write out your answers in the body text (were relevant). When you have completed your assignment, click the "Knit" button to render your `.RMD` file into a `.HTML` report. 

A physical copy of the assignment must be turned in during class on the corresponding due date. This assignment is *DUE* on **2/7**.

<br>

### Common Knitting Issues

`.RMD` files are knit into `.HTML` and other formats procedural, or line-by-line.

* An error in code when knitting will halt the process; error messages will tell you the specific line with the error
* Certain functions like `install.packages()` or `setwd()` are bound to cause errors in knitting 
* Altering a dataset or variable in one chunk will affect their use in all later chunks
* If an object is "not found", make sure it was created or loaded with `library()` in a previous chunk

**If All Else Fails:** If you cannot determine and fix the errors in a code chunk that's preventing you from knitting your document, add `eval = FALSE` inside the brackets of `{r}` at the beginning of a chunk to ensure that R does not attempt to evaluate it, that is: `{r eval = FALSE}`. This will prevent an erroneous chunk of code from halting the knitting process.  